---
# YAML header created by ox-ravel
title: Displaying Time Series with R
author: Oscar Perpiñán Lamigueiro
date: March 2020
---


# Introduction

A time series is a sequence of observations registered at consecutive
time instants. The visualization of time series is intended to reveal
changes of one or more quantitative variables through time, and to
display the relationships between the variables and their evolution
through time. The standard time series graph displays the time along
the horizontal axis. On the other hand, time can be conceived as a
grouping or conditioning variable. This solution allows several
variables to be displayed together with a scatterplot, using different
panels for subsets of the data (time as a conditioning variable) or
using different attributes for groups of the data (time as a grouping
variable). Finally, time can be used as a complementary variable that
adds information to a graph where several variables are confronted.

Next sections provide a variety of examples to illustrate a set of
useful techniques working with three datasets available at the `data`
folder of the repository.

This document has been prepared with content extracted and adapted
from the second edition of the book [&ldquo;Displaying time series, spatial
and space-time data with R&rdquo;](https://oscarperpinan.github.io/bookvis/) published with [Chapman&Hall/CRC](https://www.crcpress.com/Displaying-Time-Series-Spatial-and-Space-Time-Data-with-R-Second-Edition/Lamigueiro/p/book/9781138089983).

The most relevant packages are: `lattice`, `latticeExtra`, and
 `ggplot2` for static graphics; `zoo` and `xts` for reading and
 arranging data as time series; `RColorBrewer` for defining color
 palettes; and packages `dygraphs`, `highcharter`, and `plotly`, based
 on the `htmlwidgets` framework, to generate animations or interactive
 graphics.

Let&rsquo;s begin loading the main packages and setting some graphical
parameters of `lattice` and `latticeExtra`.

```{r   }
library(lattice)
library(ggplot2)
## latticeExtra must be loaded after ggplot2 to prevent masking of its
## `layer` function.
library(latticeExtra)
library(RColorBrewer)
## lattice and latticeExtra configuration
source("latticeConfig.R")

library(zoo) 
```


# Time on the Horizontal Axis

The most frequent visualization method of a time series uses the
horizontal axis to depict the time index. This section illustrates
two variants of this approach to display multivariate time series:
multiple time series with different scales, and variables with the
same scale. 


## Time Graph of Variables with Different Scales

There is a variety of scientific research interested in the
relationship among several meteorological variables. A suitable
approach is to display the time evolution of all of them using a
panel for each of the variables. The superposition of variables
with different characteristics is not very useful (unless their
values were previously rescaled), so this approach is postponed for
next section (variables with the same scale).

For the first example we will use the eight years of daily data from
a meteorological station located at Aranjuez (Madrid).  

 ```{r   }
 load('data/aranjuez.RData') 
 ```

This multivariate time series can be displayed with the `xyplot`
 method of `lattice` for `zoo` objects with a panel for each variable.

 ```{r   }
 ## The layout argument arranges panels in rows
 xyplot(aranjuez, layout = c(1, ncol(aranjuez))) 
 ```

The package `ggplot2` provides the generic method `autoplot` to
automate the display of certain classes with a simple command. The
package `zoo` provides an `autoplot` method for the `zoo` class with
a result similar to that obtained with `xyplot`.

 ```{r   }
 ## facet_free allows each panel to have its own range
 autoplot(aranjuez) + facet_free() 
 ```


## Time Series of Variables with the Same Scale

As an example of time series of variables with the same scale, we will
use measurements of solar radiation from different meteorological
stations.

 ```{r   }
 load('data/navarra.RData') 
 ```

The first attempt to display this multivariate time series makes use
of the `xyplot.zoo` method. The objective of this graphic is to
display the behavior of the collection as a whole: the series are
superposed in the same panel (`superpose=TRUE`) without legend
(`auto.key=FALSE`), using thin lines and partial
transparency. Transparency softens overplotting problems and reveals
density clusters because regions with more overlapping lines are
darker. Next code displays the variations around the time average
(`avRad`).

 ```{r   }
 avRad <- zoo(rowMeans(navarra, na.rm = 1), index(navarra))
 pNavarra <- xyplot(navarra - avRad,
		    superpose = TRUE, auto.key = FALSE,
		    lwd = 0.5, alpha = 0.3, col = 'midnightblue') 
 pNavarra 
 ```

 This result can be improved with the horizon graph. The horizon graph
is useful in examining how a large number of series changes over time,
and does so in a way that allows both comparisons between the
individual time series and independent analysis of each
series. Moreover, extraordinary behaviors and predominant patterns are
easily distinguished.

This graph displays several stacked series collapsing the y-axis
to free vertical space:

-   Positive and negative values share the same vertical
    space. Negative values are inverted and placed above the
    reference line. Sign is encoded using different hues (positive
    values in blue and negative values in red).
-   Differences in magnitude are displayed as differences in color
    intensity (darker colors for greater differences).
-   The color bands share the same baseline and are superposed, with
    darker bands in front of the lighter ones.

Because the panels share the same design structure, once this
technique is understood, it is easy to establish comparisons or spot
extraordinary events.  This method is what Tufte described as small
multiples.

Next code displays the variations of solar radiation around the time
average with an horizon graph using a row for each time series. In the
code we choose `origin=0` and leave the argument `horizonscale`
undefined (default). With this combination each panel has different
scales and the colors in each panel represent deviations from the
origin. This is depicted in the color key with the \(\Delta_i\) symbol,
where the subscript `i` denotes the existence of multiple panels with
different scales.

```{r   }
horizonplot(navarra - avRad,
            layout = c(1, ncol(navarra)),
            origin = 0, ## Deviations in each panel are calculated
                        ## from this value
            colorkey = TRUE,
            col.regions = brewer.pal(6, "RdBu")) 
```

The horizon graph is also useful in revealing the differences between
a univariate time series and another reference. For example, we might
be interested in the departure of the observed temperature from the
long-term average, or in other words, the temperature change over
time. Let&rsquo;s illustrate this approach with the time series of daily
average temperatures measured at the meteorological station of
Aranjuez. The reference is the long-term daily average calculated with
`ave`.

```{r   }
Ta <- aranjuez$TempAvg
timeIndex <- index(aranjuez)
longTa <- ave(Ta, format(timeIndex, '%j'))
diffTa <- (Ta - longTa) 
```

The next code uses `horizonplot` with the cut-and-stack method to distinguish between years. 

```{r   }
years <- unique(format(timeIndex, '%Y'))
  
horizonplot(diffTa, cut = list(n = 8, overlap = 0),
            colorkey = TRUE, layout = c(1, 8),
            scales = list(draw = FALSE, y = list(relation = 'same')),
            origin = 0, strip.left = FALSE) +
    layer(grid.text(years[panel.number()], x  =  0, y  =  0.1, 
                    gp = gpar(cex = 0.8),
                    just = "left")) 
```

An alternative is a level plot displaying the time series using parts
of its time index both as independent and as conditioning variable.
The following code displays the differences with the day of month on
the horizontal axis and the year on the vertical axis, with a
different panel for each month number. Therefore, each cell of next
figure corresponds to a certain day of the time series.

```{r   }
year <- function(x)as.numeric(format(x, '%Y'))
day <- function(x)as.numeric(format(x, '%d'))
month <- function(x)as.numeric(format(x, '%m')) 
```

```{r   }
myTheme <- modifyList(custom.theme(region = brewer.pal(9, 'RdBu')),
                      list(
                          strip.background = list(col = 'gray'),
                          panel.background = list(col = 'gray')))

maxZ <- max(abs(diffTa))

levelplot(diffTa ~ day(timeIndex) * year(timeIndex) | factor(month(timeIndex)),
          at = pretty(c(-maxZ, maxZ), n = 8),
          colorkey = list(height = 0.3),
          layout = c(1, 12), strip = FALSE, strip.left = TRUE,
          xlab = 'Day', ylab = 'Month', 
          par.settings = myTheme)
 
```

The `ggplot` version requires a `data.frame` with the day, year, and month arranged in different columns.

```{r   }
df <- data.frame(Vals = diffTa,
                 Day = day(timeIndex),
                 Year = year(timeIndex),
                 Month = month(timeIndex)) 
```

The values (`Vals` column of this `data.frame`) are displayed as a level plot thanks to the `geom_raster` function.

```{r   }
library(scales) 
## The packages scales is needed for the pretty_breaks function.

ggplot(data = df,
       aes(fill = Vals,
           x = Day,
           y = Year)) +
    facet_wrap(~ Month, ncol = 1, strip.position = 'left') +
    scale_y_continuous(breaks = pretty_breaks()) + 
    scale_fill_distiller(palette = 'RdBu', direction = 1) + 
    geom_raster() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) 
```


## Interactive Graphics

This section describes the interactive alternatives of the static
figures included in the previous sections with several packages:
`dygraphs`, `highcharter`, and `plotly`. These packages
are `R` interfaces to JavaScript libraries based on the `htmlwidgets`
package.


### Dygraphs

The `dygraphs` package is an interface to the `dygraphs` JavaScript
library, and provides facilities for charting time-series. It works
automatically with `xts` time series objects, or with objects than can
be coerced to this class. The result is an interactive graph, where
values are displayed according to the mouse position over the time
series. Regions can be selected to zoom into a time period. 

 ```{r   }
 library(dygraphs)

 dyTemp <- dygraph(aranjuez[, c("TempMin", "TempAvg", "TempMax")],
		   main = "Temperature in Aranjuez",
		   ylab = "ºC")

 dyTemp 
 ```

You can customize `dygraphs` by piping additional commands onto the
original graphic.  The function `dyOptions` provides several choices
for the graphic, and the function `dyHighlight` configures options for
data series mouse-over highlighting. For example, with the next code
the semi-transparency value of the non-selected lines is reduced and
the width of selected line is increased.

 ```{r   }
 dyTemp %>%
     dyHighlight(highlightSeriesBackgroundAlpha = 0.2,
		 highlightSeriesOpts = list(strokeWidth = 2)) 
 ```

An alternative approach to depict the upper and lower variables of
this time series is with a shaded region. The `dySeries` function
accepts a character vector of length 3 that specifies a set of input
column names to use as the lower, value, and upper for a series with a
shaded region around it.

 ```{r   }
 dygraph(aranjuez[, c("TempMin", "TempAvg", "TempMax")],
	 main = "Temperature in Aranjuez",
	 ylab = "ºC") %>%
     dySeries(c("TempMin", "TempAvg", "TempMax"),
	      label = "Temperature") 
 ```


### Highcharter

The `highcharter` package is an interface to the `highcharts`
JavaScript library, with a wide spectrum of graphics
solutions. Displaying time series with this package can be achieved
with the combination of the generic `highchart` function and several
calls to the `hc_add_series_xts` function through the pipe `%>%`
operator.  Once again, the result is an interactive graph with
selection and zoom capabilities. 

 ```{r   }
 library(highcharter)
 library(xts)

 aranjuezXTS <- as.xts(aranjuez)

 highchart() %>%
     hc_add_series(name = 'TempMax',
		       aranjuezXTS[, "TempMax"]) %>%
     hc_add_series(name = 'TempMin',
		       aranjuezXTS[, "TempMin"]) %>%
     hc_add_series(name = 'TempAvg',
		       aranjuezXTS[, "TempAvg"]) 
 ```


### plotly

The `plotly` package is an interface to the `plotly` JavaScript
library, also with a wide spectrum of graphics solutions. This package
does not provide any function specifically focused on time
series. Thus, the time series object has to be transformed in a
`data.frame` including a column for the time index. If the
`data.frame` is in *wide* format (one column per variable), each
variable will be represented with a call to the `add_lines`
function. However, if the `data.frame` is in *long* format (a column
for values, and a column for variable names) only one call to
`add_lines` is required. The next code follows this approach using the
combination of `fortify`, to convert the `zoo` object into a
`data.frame`, and `melt`, to transform from wide to long format.

 ```{r   }
 aranjuezDF <- fortify(aranjuez[,
				c("TempMax",
				  "TempAvg",
				  "TempMin")],
		       melt = TRUE)

 summary(aranjuezDF) 
 ```

Next code produces an interactive graphic with the generic function
`plot_ly` connected with `add_lines` through the pipe operator,
`%>%`.

 ```{r   }
 library(plotly)

 plot_ly(aranjuezDF) %>%
     add_lines(x = ~ Index,
	       y = ~ Value,
	       color = ~ Series)

 ```


# Time as a Conditioning or Grouping Variable

Previously we learned to display the time evolution of multiple time
  series with different scales. But, what if instead of displaying the
  time evolution we want to represent the relation between the
  variables? This section follows this approach: first, a scatterplot
  matrix using groups is defined according to the time as a grouping
  variable; next, an enhanced scatterplot with time as a conditioning
  variable is produced using the small multiples technique.


## Scatterplot Matrix: Time as a Grouping Variable

The scatterplot matrices are based on the technique of small
multiples: small, thumbnail-sized representations of multiple images
displayed all at once, which allows the reader to immediately, and in
parallel, compare the inter-frame differences.  A scatterplot matrix
is a display of all pairwise bivariate scatterplots arranged in a \(p
 \times p\) matrix for \(p\) variables. Each subplot shows the relation
between the pair of variables at the intersection of the row and
column indicated by the variable names in the diagonal panels.

This graphical tool is implemented in the `splom` function. The
following code displays the relation between the set of
meteorological variables using a sequential palette from the
ColorBrewer catalog (`RbBu`, with black added to complete a
twelve-color palette) to encode the month. The order of colors of
this palette is chosen in order to display summer months with
intense colors and to distinguish between the first and second
half of the year with red and blue, respectively.

 ```{r   }
 load('data/aranjuez.RData')
 aranjuezDF <- as.data.frame(aranjuez)
 aranjuezDF$Month <- format(index(aranjuez), '%m') 
 ```

 ```{r   }
 ## Red-Blue palette with black added (12 colors)
 colors <- c(brewer.pal(n = 11, 'RdBu'), '#000000')
 ## Rearrange according to months (darkest for summer)
 colors <- colors[c(6:1, 12:7)] 
 ```

 ```{r   }
 splom(~ aranjuezDF[1:10], ## Do not include "Month"
       groups = aranjuezDF$Month,
       auto.key = list(space = 'right', 
		     title = 'Month', cex.title = 1),
       pscale = 0, varname.cex = 0.7, xlab = '',
       par.settings = custom.theme(symbol = colors,
				 pch = 19),
       cex = 0.3, alpha = 0.1)

 ```

The `ggplot2` version of this graphic is produced thanks to the
`ggpairs` function provided by the `GGally` package.

 ```{r  ,warning=FALSE, message=FALSE }
 library(GGally)

 ggpairs(aranjuezDF,
	 columns = 1:10, ## Do not include "Month"
	 upper = list(continuous = "points"),
	 mapping = aes(colour = Month, alpha = 0.1)) 
 ```


## Scatterplot with Time as a Conditioning Variable

Previous graphics use colors to encode months. Instead, we will now
display separate scatterplots with a panel for each month. In
addition, the statistic type (average, maximum, minimum) is included
as an additional conditioning variable.

First, the dataset must be reshaped from the wide format
(one colum for each variable) to the long format (only one column for
the temperature values with one row for each observation). This task
is easily accomplished with the `melt` function included in the
`reshape2` package.

```{r   }
library(reshape2)

aranjuezRshp <- melt(aranjuezDF,
                     measure.vars = c('TempMax',
                                      'TempAvg',
                                      'TempMin'),
                     variable.name = 'Statistic',
                     value.name = 'Temperature')
 
```

This matrix of panels can be displayed with `ggplot` using
`facet_grid`. Next code uses partial transparency to cope with
overplotting, small horizontal and vertical segments (`geom_rug`) to
display points density on both variables, and a smooth line in each
panel.

 ```{r  ,warning=FALSE, message=FALSE }
 ggplot(data = aranjuezRshp, aes(Radiation, Temperature)) +
     facet_grid(Statistic ~ Month) +
     geom_point(col = 'skyblue4', pch = 19, cex = 0.5, alpha = 0.3) +
     geom_rug() +
     stat_smooth(se = FALSE, method = 'loess',
		 col = 'indianred1', lwd = 1.2) +
     theme_bw() 
 ```

The version with `lattice` needs the `useOuterStrips` function from
the `latticeExtra` package, which prints the names of the conditioning
variables on the top and left outer margins.

 ```{r   }
 useOuterStrips(
     xyplot(Temperature ~ Radiation | Month * Statistic,
	    data = aranjuezRshp,
	    between = list(x = 0),
	    col = 'skyblue4', pch = 19,
	    cex = 0.5, alpha = 0.3)) +
     layer({
	 panel.rug(..., col.line = 'indianred1',
		   end = 0.05, alpha = 0.6)
	 panel.loess(..., col = 'indianred1',
		     lwd = 1.5, alpha = 1)
     }) 
 ```


# Time as a Complementary Variable

In this section, time will be used as a complementary variable which
adds information to a graph where several variables are
confronted. We will illustrate this approach with the evolution of
the relationship between Gross National Income (GNI) and carbon
dioxide (\(CO_2\)) emissions for a set of countries extracted from the
database of the World Bank Open Data. We will try several solutions
to display the relationship between \(CO_2\) emissions and GNI over
the years using time as a complementary variable.


## Polylines

Our first approach is to display the entire data in a panel with a
scatterplot using country names as the grouping factor. Points of each
country are connected with polylines to reveal the time evolution.

 ```{r   }
 load('data/CO2.RData') 
 ```

 ```{r   }
 ## lattice version
 xyplot(GNI.capita  ~ CO2.capita, data = CO2data,
	xlab = "Carbon dioxide emissions (metric tons per capita)",
	ylab = "GNI per capita, PPP (current international $)",
	groups = Country.Name, type = 'b') 
 ```

Three improvements can be added to this graphical result: 

1.  Define a better palette to enhance visual discrimination between
    countries.
2.  Display time information with labels to show year values.
3.  Label each polyline with the country name instead of a legend.


### Choosing Colors

The `Country.Name` categorical variable will be encoded with a
qualitative palette, namely the first five colors of `Set1` palette
from the `RColorBrewer` package. Because there are more countries
than colors, we have to repeat some colors to complete the number of
levels of the variable `Country.Name`. The result is a palette with
non-unique colors, and thus some countries will share the same
color. This is not a problem because the curves will be labeled, and
countries with the same color will be displayed at enough distance.

 ```{r   }
 nCountries <- nlevels(CO2data$Country.Name)
 pal <- brewer.pal(n = 5, 'Set1')
 pal <- rep(pal, length = nCountries) 
 ```

Adjacent colors of this palette are chosen to be easily
distinguishable. Therefore, the connection between colors and
countries must be in such a way that nearby lines are encoded with
adjacent colors of the palette. A simple approach is to calculate the
annual average of the variable to be represented along the x-axis
(`CO2.capita`), and extract colors from the palette according to the
order of this value.

 ```{r   }
 ## Rank of average values of CO2 per capita
 CO2mean <- aggregate(CO2.capita ~ Country.Name,
		      data = CO2data, FUN = mean)
 palOrdered <- pal[rank(CO2mean$CO2.capita)]  

 ## simpleTheme encapsulates the palette in a new theme for xyplot
 myTheme <- simpleTheme(pch = 19, cex = 0.6, col = palOrdered) 
 ```

 ```{r   }
 ## lattice version
 pCO2.capita <- xyplot(GNI.capita  ~ CO2.capita,
		       data = CO2data,
		       xlab = "Carbon dioxide emissions (metric tons per capita)",
		       ylab = "GNI per capita, PPP (current international $)",
		       groups = Country.Name,
		       par.settings = myTheme,
		       type = 'b')

 pCO2.capita 
 ```

 ```{r   }
 ## ggplot2 version
 gCO2.capita <- ggplot(data = CO2data,
		       aes(x = CO2.capita,
			   y = GNI.capita,
			   color = Country.Name)) +
     geom_point() + geom_path() +
     scale_color_manual(values = palOrdered, guide = FALSE) +
     xlab('CO2 emissions (metric tons per capita)') +
     ylab('GNI per capita, PPP (current international $)') +
     theme_bw() 
 ```


### Labels to Show Time Information

This result can be improved with labels displaying the years to show
the time evolution. The panel function `panel.text` prints the
year labels with the combination of `+.trellis`, `glayer_` and
`panel.text`. Using `glayer_` instead of `glayer` we ensure that the
labels are printed below the lines.

 ```{r   }
 ## lattice version
 pCO2.capita <- pCO2.capita +
     glayer_(panel.text(...,
			labels = CO2data$Year[subscripts],
			  pos = 2, cex = 0.5, col = 'gray'))
 pCO2.capita 
 ```

 ```{r   }
 ## ggplot2 version
 gCO2.capita <- gCO2.capita + geom_text(aes(label = Year),
					colour = 'gray',
					size = 2.5,
					hjust = 0, vjust = 0)

 ```


### Country Names: Positioning Labels

The common solution to link each curve with the group value is to add
a legend. However, a legend can be confusing with too many items. In
addition, the reader must carry out a complex task: Choose the line,
memorize its color, search for it in the legend, and read the country
name.

A better approach is to label each line using nearby text with the
same color encoding. A suitable method is to place the labels
avoiding the overlapping between labels and lines. The package
`directlabels` includes a wide repertory of positioning methods to
cope with overlapping. The main function, `direct.label`, is able to
determine a suitable method for each plot, although the user can
choose a different method from the collection or even define a custom
method. For the `pCO2.capita` object, the best results are obtained
with `extreme.grid`.

 ```{r   }
 library(directlabels)

 ## lattice version
 direct.label(pCO2.capita,
	      method = 'extreme.grid') 
 ```

 ```{r   }
 ## ggplot2 version
 direct.label(gCO2.capita, method = 'extreme.grid') 
 ```


## Interactive Graphics: Animation

This section describes how to display the data through animation with
interactive functionalities with a solution that resembles the motion
chart product published by Gapminder.

Gapminder is an independent foundation based in Stockholm,
Sweden.  Its mission is &ldquo;to debunk devastating myths about the world
by offering free access to a fact-based world view.&rdquo; They provide free
online tools, data, and videos &ldquo;to better understand the changing
world.&rdquo; The initial development of Gapminder was the Trendalyzer
software, used by Hans Rosling in several sequences of his documentary
&ldquo;The Joy of Stats.&rdquo;

The information visualization technique used by Trendalyzer is an
interactive bubble chart. By default it shows five variables: two
numeric variables on the vertical and horizontal axes, bubble size and
color, and a time variable that may be manipulated with a slider. The
software uses brushing and linking techniques for displaying the
numeric value of a highlighted country.

We will mimic the Trendalyzer/Motion Chart solution with the package
`plotly`, using traveling bubbles of different colors and with radius
proportional to the values of the variable `CO2.PPP`. Previously, you
should watch the magistral video &ldquo;[200 Countries, 200 Years, 4 Minutes](https://www.gapminder.org/videos/200-years-that-changed-the-world-bbc/)&rdquo;.

The `plotly` package has already been used previously to create an
interactive graphic representing time in the x-axis. In this section
this package produces an animation piping the result of the `plot_ly`
and `add_markers` functions through the `animation_slider` function.

Variables `CO2.capita` and `GNI.capita` are represented in the x-axis
and y-axis, respectively.

 ```{r   }
 p <- plot_ly(CO2data,
	      x = ~CO2.capita,
	      y = ~GNI.capita,
	      sizes = c(10, 100),
	      marker = list(opacity = 0.7,
			    sizemode = 'diameter')) 
 ```

`CO2.PPP` is encoded with the circle sizes, while `Country.Name` is
represented both with colours and with labels.

 ```{r   }
 p <- add_markers(p,
		  size = ~CO2.PPP,
		  color = ~Country.Name,
		  text = ~Country.Name, hoverinfo = "text",
		  ids = ~Country.Name,
		  frame = ~Year,
		  showlegend = FALSE) 
 ```

Finally, animation is created with `animation_opts`, to customize the
frame and transition times, and with `animation_slider` to define the
slider. 

 ```{r  ,warning=FALSE, message=FALSE }
 p <- animation_opts(p,
		     frame = 1000,
		     transition = 800,
		     redraw = FALSE)

 p <- animation_slider(p,
		       currentvalue = list(prefix = "Year "))

 p 
 ```

